---
title: "Chapter 13: Relational Data"
output: html_notebook
---

## nycflights13

```{r}
library(tidyverse)
library(nycflights13)
```

### Exercises

1.  Imagine you wanted to draw (approximately) the route each plane flies from its origin to its destination. What variables would you need? What tables would you need to combine?

    ```{r}
    data(flights)
    data(airports)

    flight.paths <- 
      left_join(flights, airports,
                by = c("origin" = "faa"))
    ```

2.  I forgot to draw the relationship between `weather` and `airports`. What is the relationship and how should it appear in the diagram?

3.  `weather` only contains information for the origin (NYC) airports. If it contained weather records for all airports in the USA, what additional relation would it define with `flights`?

4.  We know that some days of the year are "special", and fewer people than usual fly on them. How might you represent that data as a data frame? What would be the primary keys of that table? How would it connect to the existing tables?
