---
title: "Chapter 25: Many Models"
output: html_notebook
---

```{r}
library(pacman)
p_load(modelr, tidyverse)
```

## gapminder

```{r}
p_load(gapminder)
gapminder
```

We want to answer: "how does life expectancy change over time for each country?"

```{r}
ggplot(gapminder, aes(year, lifeExp, group = country)) +
  geom_line(alpha = 1/3)
```

To identify the countries that don't have upward trends, we'll fit a model with a linear trend. The model will capture steady growth over time, and the residuals will show what's left.

```{r}
nz <- filter(gapminder, country == "New Zealand")

nz %>% 
  ggplot(aes(year, lifeExp)) +
  geom_line() +
  ggtitle("Full data = ")

nz_mod <- lm(lifeExp ~ year, data = nz)
nz %>% 
  add_predictions(nz_mod) %>% 
  ggplot(aes(year, pred)) +
  geom_line() +
  ggtitle("Linear trend + ")

nz %>% 
  add_residuals(nz_mod) %>% 
  ggplot(aes(year, resid)) +
  geom_hline(yintercept = 0, color = "white", size = 3) +
  geom_line() +
  ggtitle("Remaining pattern")
```

How can we easily fit this to every country?

### Nested data

Extract out the common code with a function and repeat using a map function from purrr. This time, instead of repeating an action for each variable, we want to repeat an action for each country, a subset of rows. For that, we need a nested data frame.

```{r}
by_country <- gapminder %>% 
  group_by(country, continent) %>% 
  nest()

by_country
```

```{r}
by_country$data[[1]]
```

### List-columns

With the nested data frame, we're in a good position to fit some models.

```{r}
country_model <- function(df) {
  lm(lifeExp ~ year, data = df)
}
```

We want to apply this to every data frame.

```{r}
models <- map(by_country$data, country_model)
```

Rather than storing the results in their own object, It's better to store them in as a column in the `by_country` data frame.

```{r}
by_country <- by_country %>% 
  mutate(model = map(data, country_model))
by_country
```

### Unnesting

To get the residuals, we need to call `add_residuals()` with each model-data pair.

```{r}
by_country <- by_country %>% 
  mutate(resids = map2(data, model, add_residuals))
by_country
```

Now, we can turn the list of data frames back into a regular data frame.

```{r}
resids <- unnest(by_country, resids)
resids
```

Now we can plot the residuals:

```{r}
resids %>% 
  ggplot(aes(year, resid)) +
  geom_line(aes(group = country), alpha = 1/3) +
  geom_smooth(se = FALSE)
```

And facet by continent:

```{r}
resids %>% 
  ggplot(aes(year, resid)) +
  geom_line(aes(group = country), alpha = 1/3) +
  geom_smooth(se = FALSE) +
  facet_wrap(~ continent)
```

### Model quality

```{r}
p_load(broom)
broom::glance(nz_mod)
```

We can use `mutate()` and `unnest()` to create a data frame with a row for each country:

```{r}
by_country %>% 
  mutate(glance = map(model, glance)) %>% 
  unnest(glance)
```

To drop the list columns, we use `.drop = TRUE` in `unnest()`:

```{r}
glance <- by_country %>% 
  mutate(glance = map(model, glance)) %>% 
  unnest(glance, .drop = TRUE)
glance
```

Now we can look for models that don't fit well:

```{r}
glance %>% 
  arrange(r.squared)
```

Let's plot out Africa since they have the worst models:

```{r}
glance %>% 
  ggplot(aes(continent, r.squared)) +
  geom_jitter(width = 0.5)
```

### Analyzing the Data with Multilevel Models

Multilevel models are beneficial here because of the structure of the data:

-   Time influences changes in the outcome variable for each country

-   Countries have trends that influence the outcome

-   Countries are nested within continents which can have trends

-   Finally, time-based trends can be different across countries and continents too

#### Clean the data

```{r}
library(pacman)
p_load(ggrepel, brms, cmdstanr, tidybayes, ggh4x, lme4, emmeans)

# center year on the first year of data (1952)
gapminder_clean <- gapminder %>% 
  mutate(year_c = year - 1952)

# example countries
countries <- tribble(
  ~country,       ~continent,
  "Egypt",        "Africa",
  "Sierra Leone", "Africa",
  "Pakistan",     "Asia",
  "Yemen, Rep.",  "Asia",
  "Bolivia",      "Americas",
  "Canada",       "Americas",
  "Italy",        "Europe",
  "Portugal",     "Europe"
)

gapminder_clean <- gapminder_clean %>% 
  mutate(highlight = country %in% countries$country)

glimpse(gapminder_clean)
```

#### The effect of continent, country, and time on life expectancy

```{r}
ggplot(gapminder_clean, aes(x = year, y = lifeExp, 
                      group = country, color = continent)) +
  geom_line(aes(size = highlight)) +
  geom_smooth(method = "lm", aes(color = NULL, group = NULL), 
              color = "grey60", size = 1, linetype = "21",
              se = FALSE, show.legend = FALSE) +
  geom_label_repel(data = filter(gapminder_clean, year_c == 0, highlight == TRUE),
                   aes(label = country), direction = "y", size = 3, seed = 1234,
                   show.legend = FALSE) +
  annotate(geom = "label", label = "Global trend", x = 1952, y = 50,
           size = 3, color = "grey60") +
  scale_size_manual(values = c(0.075, 1), guide = "none") +
  #scale_color_okabe_ito(order = c(2, 3, 6, 1)) +
  labs(x = NULL, y = "Life expectancy", color = "Continent") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

#### Regular regression

We start with a model that ignores continent and country-level differences.

```{r}
pooling_fit <- lm(lifeExp ~ year_c,
                  data = gapminder_clean)

tidy(pooling_fit)
```

Plot the fitted regression.

```{r}
pooling_model_preds <- gapminder_clean %>% 
  data_grid(country, year_c) %>% 
  add_predictions(pooling_fit) %>% 
  mutate(year = year_c + 1952) %>% 
  select(-year_c) %>% 
  left_join(gapminder_clean, by = c("country", "year"))

set.seed(1)
pooling_sample <- 
  pooling_model_preds %>% 
  # nest all year-country obsevations within continents 
  group_by(continent, country) %>% 
  nest() %>% 
  # randomly select all years of 2 countries from each continent
  group_by(continent) %>% 
  slice_sample(n = 2) %>% 
  # expand the data
  unnest()

# plot
ggplot(pooling_sample, aes(x = year, y = pred)) +
  # plot original data
  geom_point(aes(y = lifeExp), shape = 21) +
  # plot regression line
  geom_line(color = "red") +
  facet_nested_wrap(vars(continent, country))
```

#### Intercepts for each continent

How much does life expectancy increase on average as time passes? In the previous model, `year` captures the global trend, but it doesn't account for continent or country-specific differences. We can see the failure of this omission by visualizing the residuals:

```{r}
pooling_model_resids <- gapminder_clean %>% 
  add_residuals(pooling_fit)

set.seed(1)
pooling_resids_sample <- 
  pooling_model_resids %>% 
  # nest all year-country obsevations within continents 
  group_by(continent, country) %>% 
  nest() %>% 
  # randomly select all years of 2 countries from each continent
  group_by(continent) %>% 
  slice_sample(n = 2) %>% 
  # expand the data
  unnest()

# plot
ggplot(pooling_resids_sample, aes(x = year, y = resid)) +
  # plot residuals
  geom_line(color = "red") +
  facet_nested_wrap(vars(continent, country))
```

This plot shows the direction of our errors. Starting around 1995 in Botswana, the model predicted a much higher life expectancy than was actually the case. In general, the negative residual values mean that the model was overly optimistic - it predicted a higher life expectancy that actually occurred. The positive residual values mean that the model was too pessimistic - it predicted a lower life expectancy that actually occurred. We actually miss the mark with pretty much every country. We can do better.

```{r}
rand_int_fit <- lmer(lifeExp ~ year_c + (1 | continent),
                     data = gapminder_clean)

tidy(rand_int_fit)
```

##### Continent-level variance

The random effects show how much variation there is in life expectancy across continents:

```{r}
tidy(rand_int_fit, effects = "ran_pars")
```

##### Continent-level random effects

We can see the actual offsets with the `ranef()` function.

```{r}
continent_offsets <- ranef(rand_int_fit)$continent %>% 
  as_tibble(rownames = "continent")
continent_offsets
```

These values are the continent-specific offsets in the intercept. We can add these to the population intercept to see the actual intercept for each continent:

```{r}
coef(rand_int_fit)$continent %>% 
  as_tibble(rownames = "continent")
```

##### Visualize continent-specific trends

```{r}
rand_int_model_preds <- gapminder_clean %>% 
  data_grid(continent, country, year_c) %>% 
  add_predictions(rand_int_fit) %>% 
  mutate(year = year_c + 1952) %>% 
  select(-year_c) %>% 
  inner_join(gapminder_clean, by = c("continent", "country", "year"))

set.seed(1)
rand_int_sample <- 
  rand_int_model_preds %>% 
  # nest all year-country obsevations within continents 
  group_by(continent, country) %>% 
  nest() %>% 
  # randomly select all years of 2 countries from each continent
  group_by(continent) %>% 
  slice_sample(n = 2) %>% 
  # expand the data
  unnest()

# plot
ggplot(rand_int_sample, aes(x = year, y = pred)) +
  # plot original data
  geom_point(aes(y = lifeExp), shape = 21) +
  # plot pooling model predictions
  geom_line(data = pooling_sample, color = "red") +
  # plot regression line
  geom_line(color = "blue") +
  facet_nested_wrap(vars(continent, country))
```

How much did the continent intercept model improve the residuals?

```{r}
rand_int_model_resids <- gapminder_clean %>% 
  add_residuals(rand_int_fit)

set.seed(1)
rand_int_resids_sample <- 
  rand_int_model_resids %>% 
  # nest all year-country obsevations within continents 
  group_by(continent, country) %>% 
  nest() %>% 
  # randomly select all years of 2 countries from each continent
  group_by(continent) %>% 
  slice_sample(n = 2) %>% 
  # expand the data
  unnest()

# plot
ggplot(rand_int_resids_sample, aes(x = year, y = resid)) +
  # plot pooling model residuals
  geom_line(data = pooling_resids_sample, color = "red") +
  # plot random intercept residuals
  geom_line(color = "blue") +
  facet_nested_wrap(vars(continent, country))
```

#### Intercepts for each country

Instead of having intercepts for each continent, let's have intercepts for each country:

```{r}
int_fit <- lmer(lifeExp ~ year_c + (1 | country),
                data = gapminder_clean)
tidy(int_fit)
```

Let's examine the country offsets for our randomly sampled countries:

```{r}
ranef(int_fit)$country %>% 
  as_tibble(rownames = "country") %>% 
  filter(country %in% pooling_sample$country)
```

And the country specific intercepts:

```{r}
coef(int_fit)$country %>% 
  as_tibble(rownames = "country") %>% 
  filter(country %in% pooling_sample$country)
```

##### Visualize country-specific trends

```{r}
int_model_preds <- gapminder_clean %>% 
  data_grid(continent, country, year_c) %>% 
  add_predictions(int_fit) %>% 
  mutate(year = year_c + 1952) %>% 
  select(-year_c) %>% 
  inner_join(gapminder_clean, by = c("continent", "country", "year"))

set.seed(1)
int_sample <- 
  int_model_preds %>% 
  # nest all year-country obsevations within continents 
  group_by(continent, country) %>% 
  nest() %>% 
  # randomly select all years of 2 countries from each continent
  group_by(continent) %>% 
  slice_sample(n = 2) %>% 
  # expand the data
  unnest()

# plot
ggplot(int_sample, aes(x = year, y = pred)) +
  # plot original data
  geom_point(aes(y = lifeExp), shape = 21) +
  # plot pooling model predictions
  geom_line(data = pooling_sample, color = "red") +
  # plot countinent intercept prediction
  geom_line(data = rand_int_sample, color = "blue") +
  # plot county intercept predictions
  geom_line(color = "orange") +
  facet_nested_wrap(vars(continent, country))
```

Let's check out the residuals for this new model:

```{r}
int_model_resids <- gapminder_clean %>% 
  add_residuals(int_fit)

set.seed(1)
int_resids_sample <- 
  int_model_resids %>% 
  # nest all year-country obsevations within continents 
  group_by(continent, country) %>% 
  nest() %>% 
  # randomly select all years of 2 countries from each continent
  group_by(continent) %>% 
  slice_sample(n = 2) %>% 
  # expand the data
  unnest()

# plot
ggplot(int_resids_sample, aes(x = year, y = resid)) +
  # plot pooling model residuals
  geom_line(data = pooling_resids_sample, color = "red") +
  # plot continent intercept residuals
  geom_line(data = rand_int_resids_sample, color = "blue") +
  # plot country intercept residuals
  geom_line(color = "orange") +
  facet_nested_wrap(vars(continent, country))
```

#### Intercepts and slopes for each county + account for year-specific differences

```{r}
rand_model <- lmer(lifeExp ~ year_c + (1 | year_c) + (1 + year_c | country),
                   data = gapminder_clean)

tidy(rand_model)
```

##### Year-level random effects

```{r}
ranef(rand_model)$year_c %>% 
  as_tibble(rownames = "year") %>% 
  mutate(year = as.numeric(year) + 1952)

```

##### Visualize country-specific intercepts and slopes and year-specific intercepts

```{r}
rand_model_preds <- gapminder_clean %>% 
  data_grid(continent, country, year_c) %>% 
  add_predictions(rand_model) %>% 
  mutate(year = year_c + 1952) %>% 
  select(-year_c) %>% 
  inner_join(gapminder_clean, by = c("continent", "country", "year"))

set.seed(1)
rand_sample <- 
  rand_model_preds %>% 
  # nest all year-country obsevations within continents 
  group_by(continent, country) %>% 
  nest() %>% 
  # randomly select all years of 2 countries from each continent
  group_by(continent) %>% 
  slice_sample(n = 2) %>% 
  # expand the data
  unnest()

# plot
ggplot(rand_sample, aes(x = year, y = pred)) +
  # plot original data
  geom_point(aes(y = lifeExp), shape = 21) +
  # plot pooling model predictions
  geom_line(data = pooling_sample, color = "red") +
  # plot countinent intercept prediction
  geom_line(data = rand_int_sample, color = "blue") +
  # plot county intercept predictions
  geom_line(data = int_sample, color = "orange") +
  geom_line(color = "black") +
  facet_nested_wrap(vars(continent, country))
```

How much does this improve the residuals:

```{r}
rand_model_resids <- gapminder_clean %>% 
  add_residuals(rand_model)

set.seed(1)
rand_resids_sample <- 
  rand_model_resids %>% 
  # nest all year-country obsevations within continents 
  group_by(continent, country) %>% 
  nest() %>% 
  # randomly select all years of 2 countries from each continent
  group_by(continent) %>% 
  slice_sample(n = 2) %>% 
  # expand the data
  unnest()

# plot
ggplot(rand_resids_sample, aes(x = year, y = resid)) +
  # plot pooling model residuals
  geom_line(data = pooling_resids_sample, color = "red") +
  # plot continent intercept residuals
  geom_line(data = rand_int_resids_sample, color = "blue") +
  # plot country intercept residuals
  geom_line(data = int_resids_sample, color = "orange") +
  geom_line(color = "black") +
  facet_nested_wrap(vars(continent, country))
```

#### Full model

```{r}
full_fit <- lmer(lifeExp ~ year_c + I(year_c)^2 + (1 + year_c + I(year_c)^2 | continent / country),
                 data = gapminder_clean)
tidy(full_fit)

full_model_preds <- gapminder_clean %>% 
  add_predictions(full_fit) %>% 
  mutate(year = year_c + 1952)

set.seed(1)
full_model_sample <- 
  full_model_preds %>% 
  # nest all year-country obsevations within continents 
  group_by(continent, country) %>% 
  nest() %>% 
  # randomly select all years of 2 countries from each continent
  group_by(continent) %>% 
  slice_sample(n = 2) %>% 
  # expand the data
  unnest()

# plot
ggplot(full_model_sample, aes(x = year, y = pred)) +
  # plot original data
  geom_point(aes(y = lifeExp), shape = 21) +
  # plot pooling model predictions
  geom_line(data = pooling_sample, color = "red") +
  # plot countinent intercept prediction
  geom_line(data = rand_int_sample, color = "blue") +
  # plot county intercept predictions
  geom_line(data = int_sample, color = "orange") +
  geom_line(data = rand_sample, color = "black") +
  geom_line(color = "green") +
  facet_nested_wrap(vars(continent, country))

full_model_resids <- gapminder_clean %>% 
  add_residuals(full_fit)

set.seed(1)
full_resids_sample <- 
  full_model_resids %>% 
  # nest all year-country obsevations within continents 
  group_by(continent, country) %>% 
  nest() %>% 
  # randomly select all years of 2 countries from each continent
  group_by(continent) %>% 
  slice_sample(n = 2) %>% 
  # expand the data
  unnest()

# plot
ggplot(full_resids_sample, aes(x = year, y = resid)) +
  # plot pooling model residuals
  geom_line(data = pooling_resids_sample, color = "red") +
  # plot continent intercept residuals
  geom_line(data = rand_int_resids_sample, color = "blue") +
  # plot country intercept residuals
  geom_line(data = int_resids_sample, color = "orange") +
  geom_line(data = rand_resids_sample, color = "black") +
  geom_line(color = "green") +
  facet_nested_wrap(vars(continent, country))
```

## List-columns

## Creating list-columns

## Specifying list-columns
