---
title: "Chapter 20: Vectors"
output: html_notebook
---

```{r}
library(tidyverse)
```

## Vector basics

There are two types of vectors:

1.  Atomic vectors, of which there are six types: logical, integer, double, character, complex, and raw. Integer and double are collectively known as numeric vectors.
2.  Lists, which are sometimes called recursive vectors because lists can contain other lists.

## Important types of atomic vector

### Exercises

1.  Describe the difference between `is.finite(x)` and `!is.finite(x)`

    The first checks to see if `x` is a finite value, while the second checks if `x` is not a finite value:

    ```{r}
    x <- 15
    is.finite(x)
    !is.finite(x)
    ```

2.  Read the source code for `dplyr::near()`. How does it work?

    ```{r}
    dplyr::near
    ```

3.  A logical vector can take 3 possible values. How many possible values can an integer vector take? How many possible values can a double take? Use google to do some research.

4.  Brainstorm at least four functions that allow you to convert a double to an integer. How do they differ? Be precise.

5.  What functions from the readr package allow you to turn a string into logical, integer, and double vector?

## Using atomic vectors

### Exercises

1.  What does `mean(is.na(x))` tell you about a vector `x`? What about `sum(!is.finite(x))`?

    This would tell you the proportion of missing values in a vector `x`. The sum function would tell you the total number of non-finite values in `x`.

2.  Carefully read the documentation of `is.vector()`. What does it actually test for? Why does `is.atomic()` not agree with the definition of atomic vectors above?

3.  Compare and contrast `setNames()` with `purrr::set_names()`.

4.  Create functions that take a vector as input and returns:

    1.  the last value. Should you use `[` or `[[`?

    ```{r}
    x <- c(1:100)

    x[length(x)]

    last_value <- function(x) {
      x[length(x)]
    }
    last_value(x)
    ```

    1.  The elements at even numbered positions.

    ```{r}
    !is.na(x %% 2)

    find_even <- function(x) {
      x[x %% 2 == 0]
    }
    find_even(x)
    ```

    1.  Every element except the last value.

    ```{r}
    not_last <- function(x) {
      x[-length(x)]
    }
    not_last(x)
    ```

    1.  Only even numbers (and no missing values).

    ```{r}
    select_even <- function(x) {
      x[x %% 2 == 0]
    }
    select_even(x)
    ```

5.  Why is `x[-which(x > 0)]` not the same as `x[x<= 0]`?

    ```{r}
    x[-which(x > 5)]

    x[x <= 5]
    ```

6.  What happens when you subset with a positive integer that's bigger than the length of the vector? What happens when you subset with a name that doesn't exist?

## Recursive vectors (lists)

```{r}
a <- list(a = 1:3, b = "a string", c = pi, d = list(-1, -5))

a[[4]][[1]]
```

### Exercises

1.  Draw the following lists as nested sets:

    1.  `list(a, b, list(c, d), list(e, f))`

    2.  `list(list(list(list(list(a)))))`

2.  What happens if you subset a tibble as if you're subsetting a list? What are the key differences between a list and a tibble?

## Attributes

## Augmented vectors

### Exercises
