---
title: 'Chapter 7: Exploratory Data Analysis'
output:
  html_document:
    df_print: paged
---

# Introduction

EDA is an iterative cycle where you:

1.  Generate questions about your data.

2.  Search for answers by visualizing, transforming, and modeling your data.

3.  Use what you learn to refine your questions and/or generate new questions.

## Prerequisites

```{r}
library(tidyverse)
```

# Questions

Your goal during EDA is to develop an understanding of your data. Two types of questions will always be useful for making discoveries within your data:

1.  What type of variation occurs within my variables?
2.  What type of covariation occurs between my variables?

## Typical values

-   Which values are the most common? Why?

-   Which values are rare? Why? Does that match your expectations?

-   Can you see any unusual patterns? What might explain them?

The histogram below suggests several interesting questions:

-   Why are there more diamonds at whole carats and common fractions of carats?

-   Why are there more diamonds slightly to the right of each peak than there are slightly to the left of each peak?

-   Why are there no diamonds bigger than 3 carats?

```{r}
diamonds %>% 
  filter(carat < 3) %>% 
  ggplot(mapping = aes(x = carat)) +
  geom_histogram(binwidth = 0.01)
```

Clusters of similar values suggest that subgroups exist in your data. To understand the subgroups, ask:

-   How are the observations within each cluster similar to each other?

-   How are the observations in separate clusters different from each other?

-   How can you explain or describe the clusters?

-   Why might the appearance of clusters be misleading?

## Exercises

1.  Explore the distribution of each of the `x`, `y`, and `z` variables in `diamonds`. What do you learn? Think about a diamond and how you might decide which dimension is the length, width, and depth.

    ```{r}
    ggplot(data = diamonds, aes(x = x)) +
      geom_histogram(binwidth = 0.05)

    ggplot(data = diamonds, aes(x = y)) +
      geom_histogram(binwidth = 0.05)

    ggplot(data = diamonds, aes(x = z)) +
      geom_histogram(binwidth = 0.05)
    ```

    `x` appears to have values of 0s which would indicate some data entry error and `z` and `y` have outliers on the right hand side.

    ```{r}
    summary(select(diamonds, x, y, z))
    ```

    The values of 0 are likely errors and the max values are suspicious.

    ```{r}
    filter(diamonds, x == 0 | y == 0 | z == 0)
    ```

    We can check for outliers by plotting the relationship between two variables.

    ```{r}
    ggplot(data = diamonds, aes(x = x, y = y)) +
      geom_point()

    ggplot(data = diamonds, aes(x = x, y = z)) +
      geom_point()

    ggplot(data = diamonds, aes(x = y, y = z)) +
      geom_point()
    ```

    To figure out what `x`, `y`, and `z` measure, we could look at some summary statistics.

    ```{r}
    summarise(diamonds, x_bar = mean(x), y_bar = mean(y), z_bar = mean(z))
    ```

    `z` seems to be smaller than `x` and `y` so I would say that `z` is the depth and `x` and `y` are the length and widths.

2.  Explore the distribution of `price`. Do you discover anything unusual or surprising? (Hint: Carefully think about the `binwidth` and make sure you try a wide range of values.)

    ```{r}
    summary(diamonds$price)

    ggplot(data = diamonds, aes(x = price)) +
      geom_histogram(binwidth = 5)
    # this bin is too narrow

    ggplot(data = diamonds, aes(x = price)) +
      geom_histogram(binwidth = 5) +
      coord_cartesian(xlim = c(0, 2000))

    ggplot(data = diamonds, aes(x = price)) +
      geom_histogram(binwidth = 50) +
      coord_cartesian(xlim = c(0, 2000))
    ```

3.  How many diamonds are 0.99 carat? How many are 1 carat? What do you think is the cause of the difference?

    ```{r}
    diamonds %>% count(carat == 0.99)
    diamonds %>% count(carat == 1)

    # or
    diamonds %>% 
      filter(carat == 0.99 | carat == 1) %>% 
      count(carat)

    diamonds %>% 
      filter(carat == 0.99 | carat == 1) %>% 
      ggplot(aes(x = carat)) +
      geom_histogram()

    # or
    ggplot(data = diamonds, aes(x = carat)) +
      geom_histogram(binwidth = 0.01) +
      coord_cartesian(xlim = c(0.99, 1))
    ```

    People are more likely to buy 1 carat diamonds, so retailers round up the size.

4.  Compare and contrast `coord_cartesian()` vs `xlim()` or `ylim()` when zooming in on a histogram. What happens if you leave `binwidth` unset? What happens if you try and zoom so only half a bar shows?

    ```{r}
    ggplot(data = diamonds, aes(x = price)) +
      geom_histogram()

    ggplot(data = diamonds, aes(x = price)) +
      geom_histogram() +
      coord_cartesian(xlim = c(0, 5000))

    ggplot(data = diamonds, aes(x = price)) +
      geom_histogram() +
      xlim(0, 5000)
    ```

    `xlim()` and `ylim()` influence actions before the calculation of the stats and `coord_cartesian()` applies after the stats have been calculated.

# Missing Values

```{r}
nycflights13::flights %>% 
  mutate(cancelled = is.na(dep_time),
         sched_hour = sched_dep_time %/% 100,
         sched_min = sched_dep_time %% 100,
         sched_dep_time = sched_hour + sched_min / 60) %>% 
  ggplot(aes(x = sched_dep_time, color = cancelled)) +
  geom_freqpoly(binwidth = 0.25)
```

## Exercises

1.  What happens to missing values in a histogram? What happens to missing values in a bar chart? Why is there a difference?

    Missing values are removed from histograms.

    ```{r}
    diamonds2 <- 
      diamonds %>% 
      mutate(y = ifelse(y < 3 | y > 20, NA, y),
             cut = ifelse(cut == "Fair", NA, as.character(cut)))

    ggplot(data = diamonds2, mapping = aes(x = y)) + 
      geom_histogram()
    ```

    In a bar chart, `NA` values are considered a different category.

    ```{r}
    ggplot(data = diamonds2, aes(x = cut)) + 
      geom_bar()
    ```

2.  What does `na.rm = TRUE` do in `mean()` and `sum()`?

    `na.rm = TRUE` removes missing values before preforming the operation.

# Covariation

## A categorical and continuous variable

Instead of displaying the count on the y-axis of a `geom_frepoly()` plot, we can display the density, which is the count standardized so that the area under each frequency polygon is one.

```{r}
ggplot(data = diamonds, mapping = aes(x = price, y = ..density..)) + 
  geom_freqpoly(mapping = aes(colour = cut), binwidth = 500)
```

### Exercises

1.  Use what you've learned to improve the visualization of the departure times of cancelled vs. non-cancelled flights.

    ```{r}
    nycflights13::flights %>% 
      mutate(cancelled = is.na(dep_time),
             sched_hour = sched_dep_time %/% 100,
             sched_min = sched_dep_time %% 100,
             sched_dep_time = sched_hour + sched_min / 60) %>% 
      ggplot(aes(x = sched_dep_time, y = ..density.., color = cancelled)) +
      geom_freqpoly(binwidth = 0.25)
    ```

2.  What variable in the diamonds dataset is most important for predicting the price of a diamond? How is that variable correlated with cut? Why does the combination of those two relationships lead to lower quality diamonds being more expensive?

    ```{r}
    head(diamonds)

    cor(diamonds$price, select(diamonds, carat, x, y, z))

    ggplot(data = diamonds, aes(x = carat, y = price)) +
      geom_point()

    ggplot(data = diamonds, aes(x = carat, y = cut)) +
      geom_boxplot()

    ggplot(data = diamonds, aes(x = carat, y = price, color = cut)) +
      geom_point()
    ```

    There is a weak correlation between carat and cut. People would rather buy a larger diamond at the expense of cut quality.

3.  Install the ggstance package, and create a horizontal boxplot. How does this compare to using `coord_flip()`?

    ```{r}
    ggplot(data = diamonds, aes(x = carat, y = cut)) +
      geom_boxplot() +
      coord_flip()

    #install.packages("ggstance")
    library(ggstance)
    ggplot(data = diamonds, aes(x = carat, y = cut)) +
      geom_boxploth()
    ```

4.  One problem with boxplots is that they were developed in an era of much smaller datasets and tend to display a prohibitively large number of "outlying values". One approach to remedy this problem is the letter value plot. Install the `lvplot` package, and try using `geom_lv()` to display the distribution of price vs cut. What do you learn? How do you interpret the plots?

    ```{r}
    #install.packages("lvplot")
    library(lvplot)

    ggplot(data = diamonds, aes(x = cut, y = price)) +
      geom_lv()
    ```

5.  Compare and contrast `geom_violin()` with a facetted `geom_histogram()`, or a coloured `geom_freqpoly()`. What are the pros and cons of each method?

    ```{r}
    ggplot(data = diamonds, aes(x = cut, y = price)) +
      geom_violin()

    ggplot(data = diamonds, aes(x = price)) +
      geom_histogram() +
      facet_wrap(~ cut, ncol = 1)

    ggplot(data = diamonds, aes(x = price, color = cut)) +
      geom_freqpoly()
    ```

6.  If you have a small dataset, it's sometimes useful to use `geom_jitter()` to see the relationship between a continuous and categorical variable. The ggbeeswarm package provides a number of methods similar to`geom_jitter()`. List them and briefly describe what each one does.

## Two categorical variables

### Exercises

1.  How could you rescale the count dataset above to more clearly show the distribution of cut within color, or color within cut?

    ```{r}
    ggplot(data = diamonds, aes(x = color)) +
      geom_bar() +
      facet_wrap(~ cut, ncol = 1)
    ```

2.  Use `geom_tile()` together with dplyr to explore how average flight delays vary by destination and month of year. What makes the plot difficult to read? How could you improve it?

    ```{r}
    nycflights13::flights %>% 
      group_by(dest, month, year) %>% 
      summarise(avg_delay = mean(dep_delay, na.rm = TRUE)) %>% 
      ggplot(aes(x = factor(month), y = dest, fill = avg_delay)) +
      geom_tile()

    # removing destinations without 12 months of flights
    nycflights13::flights %>% 
      group_by(dest, month, year) %>% 
      summarise(avg_delay = mean(dep_delay, na.rm = TRUE)) %>% 
      group_by(dest) %>% 
      filter(n() == 12) %>% 
      ungroup() %>% 
      ggplot(aes(x = factor(month), y = reorder(dest, avg_delay), fill = avg_delay)) +
      geom_tile()
    ```

3.  Why is it slightly better to use `aes(x = color, y = cut)` rather than `aes(x = cut, y = color)` in the example above?

    ```{r}
    diamonds %>% 
      count(color, cut) %>%  
      ggplot(mapping = aes(x = color, y = cut)) +
        geom_tile(mapping = aes(fill = n))

    diamonds %>% 
      count(color, cut) %>%  
      ggplot(mapping = aes(x = cut, y = color)) +
        geom_tile(mapping = aes(fill = n))
    ```

## Two continuous variables

### Exercises

1.  Instead of summarizing the conditional distribution with a boxplot, you could use a frequency polygon. What do you need to consider when using`cut_width()` vs `cut_number()`? How does that impact a visualization of the 2d distribution of `carat` and `price`?

    ```{r}
    ggplot(data = diamonds, aes(x = price, color = cut_number(carat, 5))) +
      geom_freqpoly()

    ggplot(data = diamonds, 
           aes(x = price, color = cut_width(carat, width = 1, boundary = 0))) +
      geom_freqpoly()
    ```

2.  Visualize the distribution of carat, partitioned by price.

    ```{r}
    ggplot(data = diamonds, 
           aes(x = carat, color = cut_width(price, 2500, boundary = 0))) +
      geom_density()

    ggplot(data = diamonds, 
           aes(x = carat, y = cut_width(price, 2000, boundary = 0))) +
      geom_boxplot()

    ggplot(data = diamonds, aes(x = carat, y = cut_number(price, 10))) +
      geom_boxplot()
    ```

3.  How does the price distribution of very large diamonds compare to small diamonds? Is it as you expect, or does it surprise you?

    There is much less variation in the price of small diamonds compared to large diamonds.

4.  Combine two of the techniques you've learned to visualize the combined distribution of cut, carat, and price.

    ```{r}
    ggplot(data = diamonds, aes(x = carat, y = price, fill = cut)) +
      geom_hex()

    ggplot(data = diamonds, aes(x = cut_width(carat, width = 1, boundary = 0), y = price, color = cut)) +
      geom_boxplot()
    ```

5.  Two dimensional plots reveal outliers that are not visible in one dimensional plots. For example, some points in the plot below have an unusual combination of `x` and `y` values, which makes the points outliers even though their `x` and `y` values appear normal when examined separately.

    ```{r}
    ggplot(data = diamonds) +
      geom_point(mapping = aes(x = x, y = y)) +
      coord_cartesian(xlim = c(4, 11), ylim = c(4, 11))
    ```

    Why is a scatterplot a better display than a binned plot for this case?

    A binned plot could hide the outliers.

    ```{r}
    ggplot(data = diamonds) +
      geom_point(mapping = aes(x = cut_width(x, width = 1), y = cut_width(y, 1))) +
      coord_cartesian(xlim = c(4, 11), ylim = c(4, 11))
    ```
