---
title: "Chapter 3: Data visualisation"
output: html_notebook
---

# Prerequisites

```{r}
library(tidyverse)
```

# First steps

Do cars with big engines use more fuel than cars with small engines?

## The `mpg` data frame

```{r}
mpg
```

## Creating a ggplot

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy))
```

## Exercises

**1.  Run `ggplot(data = mpg)`. What do you see?**

```{r}
ggplot(data = mpg)
```

**2.  How many rows are in `mpg`? How many columns?**

```{r}
# rows
nrow(mpg)

# columns
ncol(mpg)
```

**3.  What does the `drv` variable describe? Read the help for `?mpg` to find out?**

```{r}
glimpse(mpg)
```

**4. Make a scatterplot of `hwy` vs `cyl`.**

```{r}
ggplot(data = mpg, aes(x = cyl, y = hwy)) +
  geom_point()
```

**5. What happens if you make a scatterplot of `class` vs `drv`? Why is the plot not useful?**

```{r}
ggplot(data = mpg, aes(x = class, y = drv)) +
  geom_point()
```

# Aesthetic mappings

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point()
```

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy, size = class)) +
  geom_point()
```

```{r}
# Left
ggplot(data = mpg, aes(x = displ, y = hwy, alpha = class)) +
  geom_point()

# Right
ggplot(data = mpg, aes(x = displ, y = hwy, shape = class)) +
  geom_point()
```

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point(color = "blue")
```

## Exercises

**1. What's wrong with this code? Why are the points not blue?**

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = "blue"))
```

The manual color setting needs to be outside of the `aes` argument. 

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue")
```

**2. Which variables in `mpg` are categorical? Which variables are continuous? How can you see this information when you run `mpg`?**

```{r}
glimpse(mpg)
```

**3. Map a continuous variable to `color`, `size`, and `shape`. How do these aesthetics behave differently for categorical vs. continuous variables?**

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = year))

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = year))

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = year))
```

**4. What happens if you map the same variable to multiple aesthetics?**

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = displ, size = hwy))
```

**5. What does the `stroke` aesthetic do? What shapes does it work with?**

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, stroke = 2), shape = 21)
```

**6. What happens if you map an aesthetic to something other than a variable name, like `aes(color = displ < 5)`? Note, you'll also need to specify x and y.**

```{r}
ggplot(data = mpg, 
       aes(x = displ, y = hwy, color = displ < 5)) + 
  geom_point()
```

# Common problems

# Facets

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_wrap(~ class, nrow = 2)
```

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  facet_grid(drv ~ cyl)
```

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  facet_grid(drv ~ cyl)
```

## Exercises

**1. What happens if you facet on a continuous variable?**


```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  facet_wrap(~ cty)
```

It facets along all combinations of the variable. 

**2. What do the empty cells in plot with `facet_grid(drv ~ cyl)` mean? How do they relate to this plot?**

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = drv, y = cyl))
```


```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  facet_grid(drv ~ cyl)
```

It means that there is no data in the combination of variables. 

**3. What plots does the following code make? What does the `.` do?**

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(drv ~ .)

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(. ~ cyl)
```

The `.` allows the user to specify facets by rows or columns. 

**4. Take the first faceted plot in this section:**

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2)
```

**What are the advantages to using faceting instead of the color aesthetic? What are the disadvantages? How might the balance change if you had a larger dataset?**

**5. Read `?facet_wrap`. What does `nrow` do? What does `ncol` do? What other options control the layout of the individual panels? Why doesn't `facet_grid()` have `nrow` and `ncol` arguments?**

**6. When using `facet_grid()` you should usually put the variable with more unique levels in the columns. Why?**

# Geometric objects

## Exercises

**1. What deom would you use to draw a line chart?**

`geom_line()`

**A boxplot?**

`geom_boxplot()`

**A histogram?**

`geom_histogram()`

**An area chart?**

`geom_area()`

**2. Run this code in your head and predict what the output will look like. Then, rune the code in R and check you predictions.**

This code will produce a scatterplot with a fitted line. 

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) + 
  geom_point() + 
  geom_smooth(se = FALSE)
```

**3. What does `show.legend = FALSE` do? What happens if you remove it? Why do you think I used it earlier in the chapter?**

`show_legend` supresses the legend mappings. 

**4. What does the `se` argument to `geom_smooth()` do?**

It contols the standard error shading in the plot. 

**5. Will these two graphs look different? Why/why not?**

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()

ggplot() + 
  geom_point(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))
```

No, they will produce the same plot. 

**6. Recreate the R code necessary to generate the following graphs.**

```{r}
plot1 <- 
  ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point(size = 4) +
  geom_smooth(se = FALSE)

plot2 <- 
  ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point(size = 4) +
  geom_smooth(aes(group = drv), se = FALSE)

plot3 <- 
  ggplot(data = mpg, aes(x = displ, y = hwy, color = drv)) +
  geom_point(size = 4) +
  geom_smooth(aes(group = drv), se = FALSE)

plot4 <- 
  ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point(size = 4, aes(color = drv)) +
  geom_smooth(se = FALSE)

plot5 <- 
  ggplot(data = mpg, aes(x = displ, y = hwy, color = drv)) +
  geom_point(size = 4) +
  geom_smooth(aes(linetype = drv), se = FALSE)

plot6 <- 
  ggplot(data = mpg, aes(x = displ, y = hwy, color = drv)) +
  geom_point(size = 4, color = "white") +
  geom_point()

library(patchwork)

(plot1 + plot2) / (plot3 + plot4) / (plot5 + plot6)
```

# Statistical transformations

## Exercises

**1. What is the default geom associated with `stat_summary()`? How could you rewrite the previous plot to use that geom function instead of the stat function?**

The default is `geom_pointrange()` and the default stat for this geom is `identity()`. 

```{r}
ggplot(data = diamonds) +
  geom_pointrange(aes(x = cut, y = depth), stat = "summary")
```

**3. Most geoms and stats come in pairs that are almost always used in concert. Read through the documentation and make a list for all the pairs. What do they have in common?**

**4. What variables does `stat_smooth()` compute? What parameters control its behavior?**

It computes a predicted value, a confidence interval and and standard error.

**5. In our proportion bar chart, we need to set `group = 1`. Why? In other words what is the problem with these two graphs?**

```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y = after_stat(prop)))

ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = color, y = after_stat(prop)))
```

# Position adjustments

*1. What is the problem with this plot? How could you improve it?*

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
  geom_point()
```

There are too many points that are overlapping? It could be improved by adding jitter. 

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
  geom_jitter()
```

*2. What parameters to `geom_jitter()` control the amount of jittering?*

`width` and `height`. 

*3. Compare and contrast `geom_jitter()` with `geom_count()`.*

`geom_jitter()` controls the random noise in the position of the points. `geom_count()` changes the size of the points based on the number of overlapping points. 

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
  geom_jitter()

ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
  geom_count()
```

*4. What's the default position adjustment for `geom_boxplot()`? Create a visualization of the `mpg` dataset that demonstraits it.*

The default position of `gom_boxplot()` is "dodge2" which does not change the vertical position, but changes the horizontal position to avoid having boxplots stack on each other. 

```{r}
ggplot(data = mpg, mapping = aes(x = class, y = cty, color = drv)) + 
  geom_boxplot(position = "dodge2")
```

With `position_identity()` the boxes will overlap:

```{r}
ggplot(data = mpg, mapping = aes(x = class, y = cty, color = drv)) + 
  geom_boxplot(position = "identity")
```

# Coordinate systems

*1. Turn a stacked bar chart into a pie chart using `coord_polar()`.*

```{r}
ggplot(data = mpg, mapping = aes(x = 1, fill = class)) + 
  geom_bar() +
  coord_polar(theta = "y")
```

*2. What does `labs()` do? Read the documentation.*

`labs()` adds labels to the plot. 

*3. What's the difference between `coord_quickmap()` and `coord_map()`?*

```{r}
nz <- map_data("usa")

ggplot(nz, aes(long, lat, group = group)) +
  geom_polygon(fill = "white", colour = "black") +
  coord_quickmap()

ggplot(nz, aes(long, lat, group = group)) +
  geom_polygon(fill = "white", colour = "black") +
  coord_map()
```

`coord_quickmap()` is faster but ignores the curvature of the earth. 

*4. What does the plot below tell you about the relationship between city and highway mpg? Why is `coord_fixed()` important? What does `geom_abline()` do?*

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_point() + 
  geom_abline() +
  coord_fixed()
```

This plot tells me that there is a strong, positive relationship between these variables. `geom_abline()` adds a 45 degree line to the data. 

```{r}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_point() + 
  geom_abline()
```

`coord_fixed()` maintains the 45 degree line. 

